\chapter{Methodology}\label{chapt:methodology}


\section{Research design}\label{sec:methodology:design}
This thesis is situated within the constructive research tradition of computer science: its primary contribution is a designed and built artefact rather than an empirical observation or theoretical derivation.
The researcher functions as a designer and builder throughout; the research question is answered by demonstrating that a solution can be constructed and that it satisfies stated criteria.

Design Science Research (DSR)~\cite{Hevner2004DesignScienceIS} is the governing methodology.
DSR is the established framework for research whose contribution is a novel IT artefact evaluated against a practical problem; it provides both a process model for artefact construction and an evaluative standard grounded in utility rather than optimality~\cite{Hevner2004DesignScienceIS}.
This thesis satisfies that condition directly: the contribution is a quantum-implicit programming language and its supporting formal and computational infrastructure, evaluated against the abstraction gap identified in the quantum programming language literature.

The IS Research Framework~\cite{Hevner2004DesignScienceIS} structures the relationship between the research and its broader context through two connecting cycles.
The \textit{relevance cycle} links the research to the environment that motivates it: the abstraction gap in quantum programming languages, established via the literature review and the QPL systematic review in Chapter~\ref{chapt:pub_qpl_review}.
Chapter~\ref{chapt:pub_qpl_review} defines the problem and supplies the requirements the artefact must satisfy.
The \textit{rigor cycle} links the research to the knowledge base it draws upon and contributes to: the formal methods and type theory that supply the theoretical foundations, and the systematic review methodology that governs the QPL survey.

The thesis produces artefacts across all four of the types identified by the IS Research Framework~\cite{Hevner2004DesignScienceIS}.
Table~\ref{tab:methodology:artefact-types} catalogues them.

\begin{table}[h]
    \centering
    \begin{tabular}{lp{0.65\textwidth}}
        \toprule
        \textbf{Artefact type} & \textbf{This thesis}                                                                                              \\
        \midrule
        Constructs             & The quantum-implicit definition; the abstraction spectrum; the quantum bias checklist.                            \\
        Models                 & The QPL scoring rubric; the QPL taxonomy produced by the systematic review in Chapter~\ref{chapt:pub_qpl_review}. \\
        Methods                & The QPL systematic review protocol; the compilation pipeline.                                                     \\
        Instantiation          & The quantum-implicit language and its compiler.                                                                   \\
        \bottomrule
    \end{tabular}
    \caption{Artefact types present in this thesis~\cite{Hevner2004DesignScienceIS}.}
    \label{tab:methodology:artefact-types}
\end{table}

Three alternative research designs are considered and rejected.
Empirical and experimental methods require a pre-existing system to observe or experiment upon; no such system exists prior to this work.
A proof-only approach would establish the formal properties of the type system and compilation pipeline but cannot account for the design and construction of the language and compiler as research contributions.
Qualitative methods are not applicable to artefact construction.
DSR explicitly supports the iterative design-build-evaluate cycle appropriate for language and compiler development and is therefore the appropriate choice.

The most substantive objection to the research design is that improved compilation technology renders a new source language unnecessary.
If a compiler can automatically infer quantum-level decisions, a quantum-implicit language provides no additional value.
This objection fails because a compiler can only infer what the source language provides.
Quantum-explicit languages encode quantum-level decisions at the vocabulary level; gate-level vocabulary encodes circuit structure directly, leaving no higher-level intent for a compiler to recover.
The QPL Silq~\cite{bichsel2020silq} illustrates the ceiling of this approach: automated uncomputation reduces programmer burden, yet the programmer's mental model remains quantum-explicit throughout.
Circuit-level optimisation techniques such as ZX-calculus rewriting and VOQC are valuable at the compilation layer; they do not address the source-level abstraction gap this thesis targets.

DSR evaluates artefacts against utility for the stated problem rather than against a standard of optimality~\cite{Hevner2004DesignScienceIS}.
The formal proofs in Chapters~\ref{chapt:pub_formal_theory} and~\ref{chapt:pub_compilation} establish the internal correctness of the artefact's \textit{methods}: the type system and the compilation pipeline as processes.
They do not independently establish the usability of the \textit{instantiation} or the validity of the \textit{constructs}, which are addressed separately through the evaluation criteria in Section~\ref{sec:methodology:evaluation}.


\section{DSR process mapping}\label{sec:methodology:process}
This research follows a problem-centred entry point into the DSRM process~\cite{Peffers2007DSRM}: the abstraction gap in quantum programming languages is established as the motivating problem before solution goals are defined.
The QPL systematic review in Chapter~\ref{chapt:pub_qpl_review} clarifies the solution requirements, grounding the entry point in goal-centred form.
Table~\ref{tab:methodology:dsrm-mapping} maps each DSRM activity to the corresponding thesis contribution.

\begin{table}[h]
    \centering
    \begin{tabular}{lp{0.55\textwidth}}
        \toprule
        \textbf{DSRM activity}                & \textbf{This thesis}                                                                                                                                                                                               \\
        \midrule
        Problem identification and motivation & Abstraction gap in quantum programming languages: quantum vocabulary is required by all existing languages; established via literature review and the QPL systematic review in Chapter~\ref{chapt:pub_qpl_review}. \\
        Objectives of a solution              & A quantum-implicit language: programming language abstraction model,compilable to gate-based quantum hardware, accessible to programmers without quantum expertise.                                                \\
        Design and development                & Language specification, formal theoretical basis and compilation framework in Chapters~\ref{chapt:pub_language_design},~\ref{chapt:pub_formal_theory}and~\ref{chapt:pub_compilation}.                              \\
        Demonstration                         & Proof-of-concept programs; compiler output on simulators and NISQ hardware in Chapter~\ref{chapt:pub_compilation}.                                                                                                 \\
        Evaluation                            & Formal type-safety proof; compilation correctness for a core fragment; evaluation against the success criteria defined in Section~\ref{sec:methodology:evaluation}.                                                \\
        Communication                         & Four incorporated publications and this thesis.                                                                                                                                                                    \\
        \bottomrule
    \end{tabular}
    \caption{DSRM activity mapping for this thesis~\cite{Peffers2007DSRM}.}
    \label{tab:methodology:dsrm-mapping}
\end{table}

The central research question (CRQ) and subsidiary research questions (SRQ1--SRQ3), defined in Chapter~\ref{chapt:introduction}, map directly onto this activity structure.
Table~\ref{tab:methodology:srq-mapping} shows the method, evidence and evaluation criteria corresponding to each question.

\begin{table}[h]
    \centering
    \footnotesize
    \renewcommand{\arraystretch}{1.6}
    \begin{tabular}{@{}p{0.3\textwidth}p{0.22\textwidth}p{0.22\textwidth}p{0.18\textwidth}@{}}
        \toprule
        \textbf{Question}              & \textbf{Method}                                                             & \textbf{Evidence}                                                                                       & \textbf{Criteria} \\
        \midrule
        CRQ: quantum-implicit language & DSR end-to-end; answered collectively across thesis& All artefact components& All six criteria \\
        SRQ1: Language design          & Formal language design                                                      & Language specification; formal syntax, semantics and type system& Criteria 5--6 \\
        SRQ2: Type safety              & Formal proof (progress and preservation)                                    & Type-safety proof in Chapter~\ref{chapt:pub_formal_theory}& Criterion 1 \\
        SRQ3: Compilation correctness  & Formal semantic preservation argument; end-to-end compilation demonstration& Proofs in Chapter~\ref{chapt:pub_formal_theory}; compiler output in Chapter~\ref{chapt:pub_compilation}& Criteria 2--4 \\
        \bottomrule
    \end{tabular}
    \caption{Research question to method and evidence mapping; for full question
    wording see Chapter~\ref{chapt:introduction}.}
    \label{tab:methodology:srq-mapping}
\end{table}

The formal proofs span the design and development and evaluation activities simultaneously: they constitute both a component of the design deliverable and the primary evaluation mechanism for SRQ2 and SRQ3.
This dual role reflects the nature of constructive programming language research,in which correctness is demonstrated by construction rather than by post-hoc testing.
The DSRM framework makes explicit a research design that computing research typically treats as implicit~\cite{Peffers2007DSRM}.


\section{Artefact definition and scope}\label{sec:methodology:artefact}
The artefact comprises three integrated components.
The \textit{language specification} provides a formal definition of the syntax, semantics and type system; it constitutes the primary design contribution and is the subject of Chapter~\ref{chapt:pub_language_design}.
The \textit{formal theoretical basis} establishes type safety (progress and preservation) for the core language and semantic preservation for a defined core fragment of the compilation pipeline; this is developed in Chapter~\ref{chapt:pub_formal_theory}.
The \textit{compilation framework} lowers programs in the source language to executable form on gate-based NISQ hardware; the compilation target is selected and justified in Chapter~\ref{chapt:pub_compilation}.

The thesis adopts a \textit{gate-based quantum computing} model: computation proceeds via discrete gate operations on qubits.
Measurement-based quantum computing and other execution paradigms are out of scope.
The artefact scope also excludes: hardware-specific scheduling and pulse-level control; quantum error correction at the language level; and full compiler verification, which is identified as future work.

Several design assumptions govern the artefact definition.
The gate-based execution model is assumed throughout; this boundary delimits the scope of all formal claims.
The term \textit{compilable} is operationally defined: the thesis demonstrates compilation to a quantum assembly representation with a formal correctness argument for the core fragment only.
The formal proofs establish the internal correctness of the artefact's \textit{methods}: the type system and compilation pipeline as processes.
They do not, of themselves, establish the usability of the \textit{instantiation} or the validity of the \textit{constructs} against real-world programming tasks; these are addressed through the evaluation criteria in Section~\ref{sec:methodology:evaluation}.
The specific compilation target is selected during the design phase based on fitness for purpose and reported in Chapter~\ref{chapt:pub_compilation}.


\section{Evaluation criteria and thresholds}\label{sec:methodology:evaluation}
DSR evaluates artefacts against utility for the stated problem~\cite{Hevner2004DesignScienceIS}.
Six criteria operationalise this standard for the artefact defined in Section~\ref{sec:methodology:artefact}.

\begin{enumerate}

    \item \textbf{Type safety.} A formal proof of progress and preservation~\cite{WrightFelleisen1994TypeSoundness} is produced for the core language.
    \textit{Threshold: the proof is mechanically checkable or peer-reviewed as part of the language specification publication.}

    \item \textbf{Compilation correctness.} Semantic preservation is formally established for a defined core fragment of the language; the proof technique is determined during the language formalisation phase.
    \textit{Threshold: a published formal argument covering the core fragment, with
    scope boundaries and limitations stated.}

    \item \textbf{Executability.} Executability is the primary success criterion for this thesis.
    Programs written in the language compile and execute on gate-based NISQ hardware; hardware access is provided through the University of Melbourne's membership of the IBM Quantum Network~\cite{UoMIBMQuantumHub}.
    \textit{Threshold: at minimum, demonstrated execution on a simulator; execution on NISQ hardware is the target goal.}

    \item \textbf{Expressiveness.} Representative quantum algorithms can be written in the language without resorting to escape hatches.
    \textit{Threshold: at least two non-trivial algorithms are implemented and
    compiled.}

    \item \textbf{Programming language abstraction model.} Programs can be written and reasoned about using the abstractions of programming language design: types, functions, control flow and data structures.
    This criterion concerns the vocabulary of reasoning, not the execution model: programs execute on quantum hardware, but the programmer is not required to model or engage with that hardware directly.
    Quantum-mechanical vocabulary is not required for correct use of the language in the common case.
    \textit{Threshold: representative algorithmic tasks are expressible using programming language abstractions only, without requiring quantum vocabulary.}

    \item \textbf{Absence of mandatory quantum vocabulary.} A computer scientist without a quantum mechanics background can learn and use the language from its documentation.
    \textit{Threshold: case studies and example programs demonstrate that a programmer with a classical background can engage with the language without quantum vocabulary; a formal usability study is identified as future work and is not a current deliverable of this thesis.}

\end{enumerate}


\section{Scope and limitations}\label{sec:methodology:scope}
Two limitations of the evaluation design are declared in advance.

First, the evidence basis for Criterion~6 (absence of mandatory quantum vocabulary) is informal.
Accessibility to programmers without quantum domain knowledge is demonstrated via case studies and example programs; a controlled usability study is not a deliverable of this thesis and is identified as future work.
This is a declared threat to the validity of claims concerning Criterion~6; the informal evidence basis is appropriate for a first-generation language design but does not constitute the same quality of evidence as a formal proof or peer-reviewed empirical study.

Second, execution on NISQ hardware is subject to availability and device characteristics at the time of evaluation.
Simulator execution constitutes the primary evidence base for Criterion~3 (executability); NISQ hardware execution is the target goal and is supported by the University of Melbourne's membership of the IBM Quantum Network~\cite{UoMIBMQuantumHub}.
Claims concerning hardware execution are bounded by the devices available during the evaluation period.


\section{Reproducibility}\label{sec:methodology:reproducibility}
All artefacts produced by this thesis are maintained in a public version-controlled repository: source code, language specification, formal proofs and compilation framework.
Build instructions and compilation pipeline documentation are included to allow independent reproduction of the compilation claims in Chapter~\ref{chapt:pub_compilation}.
The specific repository location and artefact licensing are reported in Chapter~\ref{chapt:pub_compilation}; at the time of submission, the repository is publicly accessible.

The formal proofs in Chapter~\ref{chapt:pub_formal_theory} are presented in a self-contained form; proof verification does not depend on access to the compilation toolchain.
