\documentclass[noarxiv]{../../latex_styles/quantum/quantumarticle}
%----- uncomment for POPL anonymous submission class ------%
%\documentclass[acmsmall,screen,review,anonymous,nonacm]{acmart}

\usepackage{biblatex}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{import}
\usepackage{longtable}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{braket}
\usepackage{amssymb}
\usepackage{mdframed}
\usepackage{threeparttable}

\input{../../preamble/macros}

\addbibresource{../../bib/references.bib}

\begin{document}

\title{From Assembly to Quantum-Implicit: Characterising the Abstraction Gap in Quantum Programming Systems}
\author{Constantinos Kavadias}
\email{ckavadias@student.unimelb.edu.au}

\author{Udaya Parampalli}
\email{udaya@unimelb.edu.au}

\author{Cezary Kaliszyk}
\email{cezary.kaliszyk@unimelb.edu.au}

\maketitle

\begin{abstract}
    Productive quantum programming requires programmers to reason explicitly about qubits, gates, measurement outcomes and uncomputation obligations, even when using the highest-level systems currently available.
    Prior taxonomy work has characterised the quantum programming landscape but has not established whether any existing system meets the requirements for a tier at which programmers write non-trivial programs without quantum-mechanical prerequisites.
    This paper reports a systematic literature review of eleven quantum programming systems assessed against a five-axis evaluation rubric spanning assembly, intermediate representation, framework, high-level and quantum-implicit tiers.
    The rubric operationalises quantum bias (the count of quantum-mechanical concepts required for typical tasks) alongside axis-level scores for mental model, primary artefact, safety, data model and hardware independence.
    No assessed system occupies the quantum-implicit tier; the four lower tiers are populated and each performs its designed function.
    A vocabulary ceiling on the mental-model axis appears across all high-level systems.
    The highest-scoring system retains a quantum bias count of three, consisting of the quantum mechanical constraints the literature identifies as irreducible at the circuit model boundary.
    The gap is structural; developing a new quantum programming language targeting this tier is justified.
\end{abstract}

\input{content}

\printbibliography

\end{document}
