\section{Methodology}\label{sec:pre_confirmation:methodology}
The research programme is conducted as Design Science Research (DSR), the established methodology for work whose primary contribution is a constructed artefact rather than an empirical observation or a theoretical derivation~\cite{Hevner2004DesignScienceIS}.
DSR supplies both a process for building the artefact and an evaluative standard grounded in utility for a practical problem rather than optimality~\cite{Hevner2004DesignScienceIS}.
The programme adopts a problem-centred entry into that process: the abstraction gap established in Section~\ref{sec:pre_confirmation:literature_review} fixes the problem the artefact must solve and the artefact is evaluated against the six criteria defined at the end of this section.

The planned contribution comprises three integrated components.
The \textit{language specification} is the primary design contribution: a formal definition of the syntax, semantics and type system of a quantum-implicit programming language.
The \textit{formal theoretical basis} establishes that the language behaves as specified, targeting type safety for the core language and semantic preservation for a defined core fragment of the compilation pipeline.
The \textit{compilation framework} lowers programs written in the source language to executable form on gate-based quantum hardware.
The programme adopts a gate-based model throughout; measurement-based quantum computing, hardware-specific scheduling, pulse-level control, language-level quantum error correction and full compiler verification are out of scope and identified as future work.

The contribution is a source language rather than an advance in compilation alone.
A compiler can only act on the intent its source language preserves, and quantum-explicit languages encode quantum-level decisions in their vocabulary, leaving no higher-level intent for a compiler to recover.
A language that expresses intent above the quantum substrate is therefore the precondition for compiler-managed quantum decisions, not an alternative to them.

DSR evaluates artefacts on their utility for the stated problem but does not define the criteria to do so; those follow from what this artefact must achieve.
The artefact must first function as a built system, with a sound type system, meaning-preserving compilation and executable output.
It must then close the abstraction gap that motivates it, expressing real algorithms through programming-language constructs without demanding quantum-mechanical knowledge of the programmer.
The six criteria below operationalise these two obligations: type safety, compilation correctness and executability test the artefact as a working system, mapping onto the formal theoretical basis and the compilation framework.
Expressiveness, the abstraction model and the absence of mandatory quantum vocabulary test whether it closes the gap, mapping onto the language specification and the requirements established in Section~\ref{sec:pre_confirmation:literature_review}.

\paragraph{Type safety} A type system is sound when no program it accepts can reach a state in which execution is undefined: a state in which evaluation has reached a term that is neither a final result nor able to take a further step, leaving the program \textit{stuck}.
A stuck program has no defined behaviour, so soundness is what allows a programmer to trust that a program the type checker accepts will not fail in an unspecified way at runtime.
The syntactic approach to type soundness establishes this guarantee through two properties~\cite{WrightFelleisen1994TypeSoundness}: \textit{progress}, that a well-typed program is either a final result or able to take a further evaluation step and \textit{preservation}, that evaluation never changes the type of a program.
Together they entail that a well-typed program never becomes stuck.
The research targets progress and preservation proofs for the core language, with peer review sought through the language specification publication.

\paragraph{Compilation correctness} Compilation is correct when the program the compiler emits means the same as the program the programmer wrote; this property is \textit{semantic preservation}, the requirement that the observable behaviour of the compiled program matches the behaviour defined by its source.
The research targets a semantic preservation argument for a defined core fragment of the pipeline, with its scope boundaries and limitations stated.

\paragraph{Executability} The language must compile to programs that run on physical hardware, not merely in principle.
The research targets compilation and execution on gate-based NISQ hardware through the University of Melbourne's membership of the IBM Quantum Network~\cite{UoMIBMQuantumHub}, with simulator execution as the minimum threshold.

\paragraph{Expressiveness} The abstraction model must be capable of expressing real quantum algorithms without recourse to lower-level quantum-explicit code; a construct that allows the programmer to bypass the high-level model and operate directly on gates is an \textit{escape hatch}.
The research targets the implementation and compilation of at least two non-trivial quantum algorithms expressed entirely within the language, without escape hatches.

\paragraph{Abstraction model} Programs must be expressible through ordinary programming-language constructs, without reference to quantum vocabulary at the surface.
The research targets a demonstration that representative algorithmic tasks are expressible through these constructs alone.

\paragraph{Absence of mandatory quantum vocabulary} A programmer with a classical background must be able to write working programs without quantum-mechanical knowledge.
The research targets case studies and example programs demonstrating this, with a formal usability study identified as future work.