\section{Research questions}\label{sec:pre_confirmation:research_questions}
The research questions structuring this thesis follow directly from the gap identified in Section~\ref{sec:pre_confirmation:literature_review}: no existing quantum programming language provides a programming model separated from the quantum substrate, and the structural constraints of quantum mechanics prevent that gap from being closed by incremental improvement to existing designs.
The central research question is:

\begin{quote}
    \textit{\textbf{CRQ: }Can a quantum-implicit programming language, whose surface model is agnostic of the underlying physical quantum substrate, be formally designed, proven correct and compiled to execute on gate-based quantum hardware?}
\end{quote}

The CRQ is decomposed into three subsidiary research questions, elements of these questions including their requirements for type safety are expanded on in Section~\ref{sec:pre_confirmation:methodology}.

\begin{quote}
    \textit{\textbf{SRQ1: }Can a quantum-implicit language be formally specified through computational abstractions requiring no knowledge of quantum mechanical principles or the circuit execution model, while remaining expressive enough to encode quantum algorithms?}
\end{quote}

\begin{quote}
    \textit{\textbf{SRQ2: }Can the core source language be proven type-safe: specifically, can \emph{progress} (well-typed programs never reach a stuck state) and \emph{preservation} (evaluation preserves the type of well-typed expressions) be formally established for the language's type system?}
\end{quote}

\begin{quote}
    \textit{\textbf{SRQ3: }Can semantic preservation be formally established for the compilation pipeline: specifically, does the compiler's translation from source language to quantum IR faithfully preserve the meaning of source programs?}
\end{quote}

SRQ1 will be addressed by the planned language specification publication, targeting the formal definition of the syntax, operational semantics and type system of the quantum-implicit language.
SRQ2 will be addressed by the planned formal theory publication, targeting progress and preservation proofs for the core language.
SRQ3 will be addressed jointly by the formal theory and compiler publications: the former will establish the semantic preservation proof and the latter will demonstrate it end-to-end for programs compiled to and executed on NISQ hardware.
The CRQ will be answered collectively across all four planned publications and synthesised in the results and evaluation chapter of the thesis.