\section{Work plan}\label{sec:pre_confirmation:work_plan}
The progress to date empirically establishes that no existing quantum programming system meets the criteria for the quantum-implicit tier.
The work of the coming three years will demonstrate that the tier is formally inhabitable by introducing a language that is specified, proven type-safe and compiled to execute on gate-based quantum hardware.

Future milestones will include the language specification, formal type system proofs and the compiler.
The language specification is the necessary first deliverable: syntax, semantics and built-in operators must be formally defined before the type system can be proven sound.
The progress and preservation proofs that follow establish semantic preservation across the compilation pipeline, providing the theoretical foundation on which the correctness of the compiler depends.

\subsection{Publication plan}\label{subsec:pre_confirmation:work_plan:publication_plan}
The research programme is structured across four publications, each addressing a distinct phase of the argument required to answer the CRQ.
The first publication, the systematic review, is drafted and targeting submission by mid-2026, it is the necessary precondition for all subsequent work.
Publication 1 establishes the empirical case for the quantum-implicit tier and defines the requirements the language design must satisfy.
The language specification follows as the highest priority planned work, targeting a draft in the 2027--2028 period.
The formal theory publication depends on the specification being settled and is planned for the 2027--2028 period with 2028--2029 as an acceptable fallback.
The compiler publication targets the 2028--2029 period and depends on the semantic preservation guarantees established by the formal theory.
If required due to time constraints, the formal theory and compiler may be developed concurrently, though the plan assumes sequential delivery.

\vspace{2em}
\begin{threeparttable}
    \begin{tabularx}{\textwidth}{@{}XXlX@{}}
        \toprule
        \textbf{Publication}                        & \textbf{Research question} & \textbf{Target year} & \textbf{Status}    \\
        \midrule
        (1) Quantum programming systematic review & CRQ (motivation)\tnote{*}           & Aug 2026             & Pending submission \\ \addlinespace
        (2) Language specification                & SRQ1, CRQ                  & 2027--2028           & Planned            \\ \addlinespace
        (3) Type system and formal theory         & SRQ2, SRQ3                 & 2027--2028           & Planned            \\ \addlinespace
        (4) Compiler                              & SRQ3, CRQ                  & 2028--2029           & Planned            \\
        \bottomrule
    \end{tabularx}
    \begin{tablenotes}
        \footnotesize
        \item[*] Targeting submission to the Journal of Systems and Software August 2026.
    \end{tablenotes}
\end{threeparttable}
\vspace{2em}

The publication sequence is constrained by pre-requisite completions for later publications.
The systematic review, completed by mid-2026, defines the requirements the language specification must satisfy.
The design decisions of the language are grounded in the empirical evidence it provides.
The language specification is the highest priority planned work, targeting a draft in the 2027--2028 period.
The formal theory publication depends on the specification being settled; progress and preservation cannot be proven for a language that is not yet defined.
The compiler publication targets the 2028--2029 period; the proven type system provides the correctness guarantees the compilation pipeline depends on.

The language specification work includes content that may warrant a separate publication.
A quantum-implicit language must define how quantum operations are expressed as language-level constructs without exposing their quantum-mechanical implementation to the programmer: superposition, entanglement, measurement and uncomputation each require a formal account at the language level.
This operations model addresses a discrete and well-scoped problem: how the compiler boundary between classical intent and quantum realisation is formally defined at the language level.
That problem is self-contained enough to constitute a standalone contribution to the programming language design literature if the material develops sufficient depth and novelty in its own right.
The decision will be made once the language specification draft is sufficiently developed to assess scope and length.
If the threshold is not met, the operations model will be incorporated directly into the language specification publication without loss to either the research programme or the formal claims of the thesis.

\subsection{Milestones and timelines}\label{subsec:pre_confirmation:work_plan:milestones}
The milestones of the research programme are presented at two levels of granularity.
The three-year candidature arc and a detailed breakdown of the tasks remaining in the 2026--2027 period.
The detailed breakdown represents the immediate preparation required before language specification work can begin.
The three-year arc is organised around three publication phases, each dependent on the one preceding it.
The 2026--2027 period has the dual obligation of completing the literature review and submitting the systematic review publication.
It also includes building the domain knowledge the language specification requires.
Preparing for the confirmation report and the language specification groundwork will occur in parallel in the transition from 2026 to 2027.

\vspace{2em}
\begin{tabularx}{\textwidth}{@{}p{2cm}lX@{}}
\toprule
\textbf{Period} & \textbf{Date} & \textbf{Milestone} \\
\midrule
\multirow{\textbf{Apr 2026--Mar 2027}}
& Aug 2026   & Systematic review submitted to peer-reviewed venue \\
& Sep 2026   & Literature review: language-level quantum operations complete \\
& Oct 2026   & Quantum operations catalogue for language specification complete \\
& Nov 2026   & Literature review: quantum compilation and IRs complete \\
& Dec 2026   & Literature review: language design complete \\
& Dec 2026   & Control flow constructs for language specification selected \\
& Feb 2027   & Confirmation written report finalised \\
& Mar 2027   & Confirmation meeting \\
\midrule
\multirow{\textbf{Apr 2027--Mar 2028}}
& Apr 2027   & Programming language theory refresher complete; language specification begins \\
& 2027--2028 & Language specification drafted and submitted \\
& 2027--2028 & Type system and formal theory work begins \\
& Mar 2028   & Quantum operations publication decision assessed \\
\midrule
\multirow{\textbf{Apr 2028--Mar 2029}}
& 2028--2029 & Compilation work drafted and submitted \\
& 2028--2029 & Formal theory completion if carried from prior period \\
& Mar 2029   & Thesis submitted \\
\bottomrule
\end{tabularx}
\vspace{2em}

The remainder of the 2026--2027 period is detailed below.
Given the proximity of these milestones to the pre-confirmation meeting, the tasks listed represent committed near-term objectives rather than indicative planning.
The period serves a dual purpose: completing the theoretical grounding required to write the language specification and fulfilling the institutional obligations of the first year of candidature.
The literature review work on quantum operations, compilation pipelines and language design theory directly informs the vocabulary and scope of the language specification.
The quantum operations catalogue and control flow selection tasks translate that theoretical grounding into concrete design inputs.
These are not separate streams but mutually reinforcing activities where insights from the literature directly shape the catalogue and vice versa.
Confirmation preparation runs concurrently with this work from January 2027, drawing on the completed literature review and the systematic review publication as its primary evidence base.

\vspace{2em}
\begin{threeparttable}
    \begin{tabularx}{\textwidth}{@{}lX@{}}
        \toprule
        \textbf{Date\tnote{*}} & \textbf{Milestone} \\
        \midrule
        Aug 2026           & Systematic review submitted to peer-reviewed venue \\
        Sep 2026           & Literature review: language-level quantum operations complete \\
        Oct 2026           & Quantum operations catalogue for language specification complete \\
        Oct 2026           & Language specification publication planned and scoped \\
        Nov 2026           & Literature review: quantum compilation and IRs complete \\
        Nov 2026           & Control flow constructs for language specification selected \\
        Dec 2026           & Literature review: language design complete \\
        Dec 2026           & Year 1 retrospective; task plan timeline updated \\
        Jan 2027           & Literature review: gap and requirements drafting begins \\
        Jan 2027           & Confirmation written report drafting begins \\
        Jan 2027           & Programming language theory refresher begins \\
        Feb 2027           & Literature review: gap and requirements complete \\
        Feb 2027           & Confirmation written report finalised \\
        Feb 2027           & iThenticate originality report obtained \\
        Feb 2027           & Confirmation seminar preparation begins \\
        Mar 2027           & Confirmation package submitted to committee \\
        Mar 2027           & Confirmation seminar rehearsals complete \\
        Mar 2027           & Confirmation meeting \\
        \bottomrule
    \end{tabularx}
    \begin{tablenotes}
        \footnotesize
        \item[*] The candidate will be overseas during August and September 2026 with limited availability. This may impact the proposed due dates for October and November 2026.
    \end{tablenotes}
\end{threeparttable}
\vspace{2em}

The milestones of the 2026--2027 period are structured to reflect the candidate's working approach to research development.
Theoretical grounding precedes design: the literature review sections on quantum operations, compilation pipelines and language design theory are completed before the language specification work begins.
The design decisions of the specification will be informed by a comprehensive understanding of the field rather than assumptions carried forward from the systematic review.
Preparatory research tasks, the quantum operations catalogue and control flow selection, run concurrently with the literature review to convert theoretical understanding into concrete design inputs.
These are not separate streams but mutually reinforcing activities where insights from the literature directly shape the catalogue and vice versa.
Publication planning for the language specification is completed in this period before any writing begins, scoping the contribution and identifying the design decisions that must be resolved before a draft can proceed.
The confirmation written report and seminar draw directly on the completed literature review and the systematic review publication.

By the end of March 2027 the research programme will have a published empirical foundation and a completed theoretical grounding across the relevant literature.
Concrete design inputs for the language specification and a scoped publication plan will also be in place.
The language specification work beginning in April 2027 proceeds from a position of deliberate and documented preparation rather than exploratory groundwork.

The research programme is governed by a structured set of recurring planning and monitoring practices.
Annual planning at the start of each candidature year translates high-level milestones into actionable work streams, with quarterly refinement of near-term tasks as the research develops.
A fortnightly literature scan using AI-assisted discovery tools maintains currency with the field, identifying work relevant to both completed sections and forthcoming research phases.
These practices ensure that progress against milestones remains visible and that the research remains responsive to developments in the quantum programming language literature.